Apache Flink\texttrademark : Streaming dataflows as a basis for universal data analytics

### 2015-11-07 paper.tex
Added some figs and listings

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Stack figure comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% This is the sutext of the figure
\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs. A Job Graph is a DAG of operators and connections between operators. There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing''). Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine. On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs. Currently these are the following: i) Gelly, an API and library for processing graphs, ii) FlinkML, an API and library for composing Machine Learning pipelines and iii) Table, an API similar in spirit to Microsoft's LINQ.

@@

\textbf{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges. There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs). For example, the graph below consists of five operators ($o_1$-$o_5$), and three intermediate results ($ir_1$-$ir_3$). Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g. reading data from a file system, a socket, a message queue, etc.), and data sinks. Operators produce intermediate results, as well as updates to state. An intermediate result is a logical handle (pointer) to the data that is produced by one operator. An intermediate result can be consumed by one or more operators. Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% JobGraph figure comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Parallelized Job Graphs.} When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph. An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys. The above JobGraph, if it is partitioned using a parallelism of two, it results to the execution graph below. Assume that $o_1$ and $o_2$ are data sources, $o_2$ is a map operator, $o_4$ is a join operator that needs both inputs partitioned by the same key, and $o_5$ is a filter. Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present). The resulting ExecutionGraph is the following:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ExecutionGraph figure comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% WindowWC snippet comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\textbf{Streaming Windows.} A window defines a logical group of records that are processed together. Flink's windowing system largely follows the Dataflow model [Dataflow] proposed by Google. A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor. 
The assigner predefines the logical groups to which each record belongs, e.g. count-based or time-based periodic windows. The trigger defines when the window operation is performed on each group. The evictor defines which records to keep in each group. The following example defines windows that operate in event time of 6 secondst hat slide every 2 seconds (Assigner). The window results are computed once the watermark passes the end of the window (Trigger).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Trigger snippet comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A short hand notation for the example above is:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Shorthand snippet comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Global window snippet comes here
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% JobGraph and Execution figs
\usepackage{tikz}
\usetikzlibrary{arrows,automata,calc,positioning,shapes.geometric}

@@

% code listings
\usepackage{listings}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

% there is no built in support or Scala yet, good enough
\lstset{frame=tb,
  language=Java,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\scriptsize\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{black},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=4,
  %for scala
  emph={%  
    object, def, val, zip%
    },emphstyle={\color{blue}}%
}%

@@

\begin{figure}[h!]
	\centering  	
  	\includegraphics[width=.45\textwidth]{figs/flink_stack.png}
	\caption{The Flink software stack.}
	\label{fig:FlinkStack}
\end{figure}

\textbf{Flink's Runtime and APIS.} The core of Flink is the distributed dataflow engine, which executes dataflow programs called Job Graphs. A Job Graph is a DAG of operators and connections between operators. There are two ``core'' APIs in Flink: the DataSet API for processing finite data sets (often referred to as ``batch processing''), and the DataStream API for processing potentially unbounded data streams (often referred to as ``stream processing''). Flink's core runtime engine can be seen as a streaming dataflow engine, and both the DataSet and DataStream APIs create programs (Job Graphs) accepted by this engine. On top of the core APIs, Flink bundles ``domain-specific'' libraries and APIs that generate DataSet and DataStream API programs. Currently these are the following: i) Gelly, an API and library for processing graphs, ii) FlinkML, an API and library for composing Machine Learning pipelines and iii) Table, an API similar in spirit to Microsoft's LINQ.

@@

\textbf{The Job Graph.} Flink's execution model is based on the Job Graph, a directed acyclic graph (DAG) that consists of nodes and edges. There are two classes of nodes: (stateful) operators, and (logical) intermediate results (IRs). For example, the graph below consists of five operators ($op_1$-$op_5$), and three intermediate results ($ir_1$-$ir_3$). Operators abstract computation (e.g., transformations, joins, etc), state (e.g., a persistent counter), as well as data sources (e.g. reading data from a file system, a socket, a message queue, etc.), and data sinks. Operators produce intermediate results, as well as updates to state. An intermediate result is a logical handle (pointer) to the data that is produced by one operator. An intermediate result can be consumed by one or more operators. Intermediate results are logical in the sense that the data they point to may or may not be materialized on disk.

\begin{figure}[ht]
\centering
\begin{tikzpicture}
[shorten >=1pt,node distance=1.75cm,>=stealth',initial text=, font=\tiny]
\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.35, draw, minimum size=1cm, text width=0.5cm, align=center]

\node[op] (op_1) {$op_1$};
\node[ir] (ir_1) [right of = op_1] {$ir_1$};
\node[op] (op_2) [right of = ir_1] {$op_2$};
\node[ir] (ir_2) [right of = op_2] {$ir_2$};
\node[op] (op_3) [below of = op_2] {$op_3$};
\node[ir] (ir_3) [right of = op_3] {$ir_3$};
\node[op] (op_4) [right of = ir_2] {$op_4$};
\node[op] (op_5) [right of = ir_3] {$op_5$};

\path[->] 	
			(op_1) 	edge node [above] {} (ir_1)
			(ir_1) 	edge node [above] {} (op_2)
			(op_2) 	edge node [above] {} (ir_2)
			(ir_2) 	edge node [above] {} (op_4)
			(op_3) 	edge node [above] {} (ir_3)
			(ir_3) 	edge node [above] {} (op_4)
			(ir_3) 	edge node [above] {} (op_5)
			;
\end{tikzpicture}
\label{fig:JobGraph}
\caption{The JobGraph is the logical view of a Flink job.}
\end{figure}

\textbf{Parallelized Job Graphs.} When the JobGraph is scheduled in a cluster for execution, it is parallelized to form an ExecutionGraph. An ExecutionGraph consists of tasks (parallel instances of operators), and intermediate result partitions (IRPs), that represent logical partitions of intermediate results corresponding to different partitioning keys. The JobGraph in fig. \ref{fig:JobGraph}, if it is partitioned using a parallelism of two, it results to the ExecutionGraph shown in fig. \ref{fig:ExecutionGraph}. Assume that $op_1$ and $op_2$ are data sources, $op_2$ is a map operator, $op_4$ is a join operator that needs both inputs partitioned by the same key, and $op_5$ is a filter. Assume also that the join is executed as a repartitioned join, i.e., both inputs are partitioned and shuffled over the cluster of 2 nodes (alternatives would be to broadcast one input or leverage a pre-existing partitioning if present). The resulting ExecutionGraph is the following:

\begin{figure}[ht]
\centering
\begin{tikzpicture}
[shorten >=1pt,node distance=1.6cm,>=stealth',initial text=, font=\tiny]
\tikzstyle{every state}=[circle, fill,draw=none,white,text=black]
\tikzstyle{op}=[circle, draw=blue!50,very thick,fill=blue!20]
\tikzstyle{ir}=[cylinder, cylinder uses custom fill, cylinder body fill=orange!50, cylinder end fill=orange!50, shape border rotate=90, aspect=0.2, draw, minimum size=0.8cm, text width=0.8cm, align=center]

\node[op] (op_11) {$op_{11}$};
\node[ir] (isp_111) [right of = op_11] {$isp_{111}$};
\node[op] (op_21) [right of = isp_111] {$op_{21}$};
\node[state] (ir_21) [right of = op_21] {};
\node[ir] (isp_212) [below of = ir_21, node distance = 0.35cm] {$isp_{212}$};
\node[ir] (isp_211) [above of = ir_21, node distance = 0.35cm] {$isp_{211}$};
\node[op] (op_31) [below of = op_21] {$op_{31}$};
\node[state] (ir_31) [right of = op_31] {};
\node[ir] (isp_312) [below of = ir_31, node distance = 0.35cm] {$isp_{312}$};
\node[ir] (isp_311) [above of = ir_31, node distance = 0.35cm] {$isp_{311}$};
\node[op] (op_41) [right of = ir_21, node distance = 2.25cm] {$op_{41}$};
\node[op] (op_51) [right of = ir_31, node distance = 2.25cm] {$op_{51}$};

\node[op] (op_22) [below of = op_31] {$op_{22}$};
\node[ir] (isp_121) [left of = op_22] {$isp_{121}$};
\node[op] (op_12) [left of = isp_121] {$op_{12}$};
\node[state] (ir_22) [right of = op_22] {};
\node[ir] (isp_222) [below of = ir_22, node distance = 0.35cm] {$isp_{222}$};
\node[ir] (isp_221) [above of = ir_22, node distance = 0.35cm] {$isp_{221}$};
\node[op] (op_32) [below of = op_22] {$op_{32}$};
\node[state] (ir_32) [right of = op_32] {};
\node[ir] (isp_322) [below of = ir_32, node distance = 0.35cm] {$isp_{322}$};
\node[ir] (isp_321) [above of = ir_32, node distance = 0.35cm] {$isp_{321}$};
\node[op] (op_42) [right of = ir_22, node distance = 2.25cm] {$op_{42}$};
\node[op] (op_52) [right of = ir_32, node distance = 2.25cm] {$op_{52}$};

\path[->] 	
			(op_11) 	edge node [above] {} (isp_111)
			(isp_111) 	edge node [above] {} (op_21)
			(op_21) 	edge node [above] {} (isp_211)
			(op_21) 	edge node [above] {} (isp_212)
			(isp_211) 	edge node [above] {} (op_41)
			(isp_212) 	edge [bend left = 12.5] node [above] {} (op_42)
			(op_31) 	edge node [above] {} (isp_311)
			(op_31) 	edge node [above] {} (isp_312)
			(isp_311) 	edge node [above] {} (op_41)			
			(isp_311) 	edge node [above] {} (op_51)
			(isp_312) 	edge node [above] {} (op_51)
			(isp_312) 	edge [bend left = 7.5] node [above] {} (op_42)

			(op_12) 	edge node [above] {} (isp_121)
			(isp_121) 	edge node [above] {} (op_22)
			(op_22) 	edge node [above] {} (isp_221)
			(op_22) 	edge node [above] {} (isp_222)
			(isp_221) 	edge [bend right = 7.5] node [above] {} (op_41)			
			(isp_222) 	edge node [above] {} (op_42)
			(op_32) 	edge node [above] {} (isp_321)
			(op_32) 	edge node [above] {} (isp_322)
			(isp_321) 	edge node [above] {} (op_52)
			(isp_321) 	edge [bend right = 12.5] node [above] {} (op_41)
			(isp_322) 	edge node [above] {} (op_52)
			(isp_322) 	edge [bend right = 7.5] node [above] {} (op_42)
			;
\end{tikzpicture}
\label{fig:ExecutionGraph}
\caption{The ExecutionGraph is the physical view of a Flink job.}
\end{figure}

@@

\lstinputlisting[label=list:WindowWordCount,caption=WindowWordCount Scala implementation,language=Java]{src/wwc.scala}

@@

\textbf{Streaming Windows.} A window defines a logical group of records that are processed together. Flink's windowing system largely follows the Dataflow model [Dataflow] proposed by Google. A window definition consists of three building blocks: a window assigner, optionally a trigger, and optionally an evictor. 
The assigner predefines the logical groups to which each record belongs, e.g. count-based or time-based periodic windows. The trigger defines when the window operation is performed on each group. The evictor defines which records to keep in each group. The following example defines windows that operate in event time of 6 secondst hat slide every 2 seconds (Assigner). The window results are computed once the watermark passes the end of the window (Trigger). Flink also offer shart hands for commonly used functionalities.

\lstinputlisting[label=list:GenWindow,caption=Window API functionalities,language=Java]{src/genwin.scala}

@@

\lstinputlisting[label=list:GlobalWindow,caption=Global Window usage,language=Java]{src/glowin.scala}
